---
title: "Opening & merging DD output views (with password protection)"
author: "Emily Urban-Wojcik"
date: "2024-05-01"
output: html_document
---

# Note
This script opens data downloaded from University of Michigan's DataDirect tool. Data download as xlsx files with multiple worksheets within one workbook. This script opens each worksheet then merges them together to make one dataframe with all downloaded variables. 

# Loading libraries/packages & paths
```{r, message = FALSE, results = 'hide', warning = FALSE}

# Un-comment and run the below lines to install these packages if you don't already have them
# install.packages("rJava")
# install.packages("XLConnect")
# install.packages("openxlsx")
# install.packages("rstudioapi") 
# install.packages("dplyr")
# install.packages("fst")

# Opening the packages we need
library(XLConnect) # package to open xlsx files with password encryption; Note that XLConnect depends on Java through the rJava. Make sure you have an up-to-date Java Development Kit (JDK) installed on. your system (not just the Java Runtime Environment). You can download the latest JDK from Oracle’s website: https://www.oracle.com/java/technologies/downloads/?er=221886.
library(openxlsx) # used to write excel files with write.xlsx
library(rstudioapi) # to ask for a password when opening xlsx file
library(dplyr) # used here for merging dataframes (full_join)
library(fst) # lightning fast serialization of data frames for R (saves data in smaller file sizes; loads more quickly)

# Setting your working directory and creating data paths; Replace with your own path; This assumes you have separate directories for raw and cleaned data within your data directory
# Note that Dropbox Teams folders are considered acceptable to use with PHI
setwd("/INSERT_YOUR_PATH_HERE/DataDirect/data/raw/")
clean_data_path <- "/INSERT_YOUR_PATH_HERE/DataDirect/data/clean/"
pw_message <- "Input password for excel sheet."

# Loading the xlsx workbook; Note that if your xlsx file is not password encrypted you can type anything into the pw box (or just remove the password call in the code below)
wb <- loadWorkbook("Example_DD_data.xlsx", password = rstudioapi::askForPassword(pw_message))

# What are the sheet names in the xlsx datafile? These will be input into later calls to open up each sheet. 
Example_data_sheets <- getSheets(wb)
print(Example_data_sheets)

```

# Opening each output view
## Demographics
```{r, message = FALSE, results = 'hide', warning = FALSE}

# Opening the worksheet
demos <- readWorksheet(wb, "DemographicInfo")

```

## Encounters
```{r, message = FALSE, results = 'hide', warning = FALSE}

# Opening the worksheet
encounters <- readWorksheet(wb, "EncounterAll")

# reformatting date in a way that is recognized by R
# AdmitDate is the date of the encounter. The date returned will be based on an algorithm that chooses the first non-null value available from the following list (in that order): HospitalAdmissionDate, CheckinTime, AppointmentTime, EncounterInstantDate, EncounterContactDate.
encounters$encounter_date <- as.POSIXct(encounters$AdmitDate, format = "%m/%d/%Y")

```

## Diagnoses
```{r, message = FALSE, results = 'hide', warning = FALSE}

# Opening the worksheets (there are two in this case with data from DiagnosesEverything; this is typically because there was too much data to fit into one worksheet for this output view for download)
diagnoses1 <- readWorksheet(wb, "DiagnosesEverything")
diagnoses2 <- readWorksheet(wb, "DiagnosesEverything (part 2)")

# Appending all dataframes starting with the same root and ending with the relevant numbers; NEED TO UPDATE THIS BASED ON NUMBER OF SHEETS
# Because there shouldn't be any overlap between a specific visit diagnosis between sheets, we append rather than merge
diagnoses <- do.call(rbind, mget(paste0("diagnoses", 1:2)))

# Checking that the number of rows matches after appending (should == TRUE); NEED TO UPDATE THIS BASED ON NUMBER OF SHEETS
nrow(diagnoses) == sum(sapply(1:2, function(i) nrow(get(paste0("diagnoses", i)))))

```

# Combining output views & saving
```{r, message = FALSE, results = 'hide', warning = FALSE}

# Merging ("joining") output views; full_join keeps all observations from each dataframe; 
# The first instance of this joins by patient, as there can be multiple encounters within each patient;
# The second instance of this joins by patient AND by encounter, as there can be multiple diagnoses within encounter
data <- full_join(demos, encounters, by = c("PatientID"))
data <- full_join(data, diagnoses, by = c("PatientID", "EncounterID"))

# Checking the number of unique patients and encounters in this worksheet
length(unique(data$PatientID))
length(unique(data$EncounterID))

# Saving as fst
write.fst(data, (paste0(clean_data_path, "Example_data_merged.fst")))

# # You can instead save as a csv file, which will then allow you to then password encrypt manually if you wish
# write.xlsx(data, ((paste0(clean_data_path, "Example_data_merged.xlsx"))))

```
