---
title: "Flagging_Diagnoses"
author: "Emily Urban-Wojcik"
date: "2025-04-16"
output: html_document
---

# Note
This script opens data from Michigan Medicine's DataDirect that were merged together in the previous script. It then flags each row for ADHD or MDD diagnoses. Then, it creates patient-level summaries of diagnoses, including first diagnosis date, number of diagnoses, and age at diagnosis.  

# Loading libraries/packages & paths
```{r, message = FALSE, results = 'hide', warning = FALSE}

# Un-comment and run the below lines to install these packages if you don't already have them
# install.packages("fst")
# install.packages("psych")
# install.packages("stringr")
# install.packages("dplyr")
# install.pakcages("tidyr")

# Opening the packages we need
library(fst) # lightning fast serialization of data frames for R (saves data in smaller file sizes; loads more quickly)
library(psych) # package for psychometric research (e.g., describe)
library(stringr) # for string operations
library(dplyr) # for data manipulation 
library(tidyr) # for pivot_wider to reshape data

# Set your working directory; Replace with your own path; This assumes you have separate directories for raw and cleaned data within your data directory
setwd("/Users/emurbanw/University of Michigan Dropbox/MED-EFDC-DDC-internal/Tools_to_Share/DataDirect/data/clean/")

# 5 patients
data <- read.fst("Example_data_merged.fst")
length(unique(data$PatientID)) 

```

# Flagging diagnoses
## ADHD diagnosis
Flagging rows as 0/1 for if they have a code for ADHD
```{r}

data$ADHD_dx <- ifelse(data$TermCodeMapped == "F90" | 
  data$TermCodeMapped == "F90.0" | 
  data$TermCodeMapped == "F90.1" | 
  data$TermCodeMapped == "F90.2" | 
  data$TermCodeMapped == "F90.8" | 
  data$TermCodeMapped == "F90.9" | 
  data$TermCodeMapped == "314.0" | 
  data$TermCodeMapped == "314.00" | 
  data$TermCodeMapped == "314.01", 1, 0)

# 8 diagnoses across 5 patients
table(data$ADHD_dx)
length(unique(data$PatientID[data$ADHD_dx==1]))

```

## Major Depressive Disorder
```{r}

##############################################################################
# MDD single episode

data$MDD_single <- ifelse(# ICD-10
  data$TermCodeMapped ==  "F32" | 
  data$TermCodeMapped ==  "F32.0" | 
  data$TermCodeMapped ==  "F32.1" | 
  data$TermCodeMapped ==  "F32.2" | 
  data$TermCodeMapped ==  "F32.3" | 
  data$TermCodeMapped ==  "F32.4" | 
  data$TermCodeMapped ==  "F32.5" | 
  data$TermCodeMapped ==  "F32.8" | 
  data$TermCodeMapped ==  "F32.89" | 
  data$TermCodeMapped ==  "F32.9" | 
  data$TermCodeMapped ==  "F32.A" | 
  # ICD-9
  data$TermCodeMapped == "296.2" | 
  data$TermCodeMapped ==  "296.20" | 
  data$TermCodeMapped ==  "296.21" | 
  data$TermCodeMapped ==  "296.22" | 
  data$TermCodeMapped ==  "296.23" | 
  data$TermCodeMapped ==  "296.24" | 
  data$TermCodeMapped ==  "296.25" | 
  data$TermCodeMapped ==  "296.26", 1, 0)

##############################################################################
# MDD recurrent episode

data$MDD_recurrent <- ifelse(# ICD-10
  data$TermCodeMapped ==  "F33" | 
  data$TermCodeMapped ==  "F33.0" | 
  data$TermCodeMapped ==  "F33.1" | 
  data$TermCodeMapped ==  "F33.2" | 
  data$TermCodeMapped ==  "F33.3" | 
  data$TermCodeMapped ==  "F33.4" | 
  data$TermCodeMapped ==  "F33.40" | 
  data$TermCodeMapped ==  "F33.41" | 
  data$TermCodeMapped ==  "F33.42" | 
  data$TermCodeMapped ==  "F33.8" | 
  data$TermCodeMapped ==  "F33.9" | 
  # ICD-9
  data$TermCodeMapped ==  "296.3" | 
  data$TermCodeMapped ==  "296.30" | 
  data$TermCodeMapped ==  "296.31" | 
  data$TermCodeMapped ==  "296.32" | 
  data$TermCodeMapped ==  "296.33" | 
  data$TermCodeMapped ==  "296.34" | 
  data$TermCodeMapped ==  "296.35" | 
  data$TermCodeMapped ==  "296.36", 1, 0)

table(data$MDD_single) # 2
table(data$MDD_recurrent) # 0

# Combining single and recurrent into one variable
data$MDD_dx <- ifelse(data$MDD_single==1 | data$MDD_recurrent==1, 1, 0)

table(data$MDD_dx) # 2

```
## Saving flagged diagnoses
```{r}

write.fst(data, "Example_data_merged_flagged.fst")

```
# Patient-Level Summary
Here we'll create a summary dataframe with just 1 row per patient
It will tell us everything we need to know for an analysis based on the long dataframe
```{r}

# Re-opening flagged data if necessary
data_flagged <- read.fst("Example_data_merged_flagged.fst")

################################################################################
# Making our summary dataframes (we will append them back together later)

# Summarizing the number of ADHD and MDD diagnoses (within patient)
summary <- data_flagged %>%
  arrange(PatientID, encounter_date) %>%
  group_by(PatientID) %>%
  summarize(num_ADHD_dx = sum(ADHD_dx, na.rm = TRUE),
            num_MDD_dx = sum(MDD_dx, na.rm = TRUE))

# Summarizing only the rows (within patient) where there's an ADHD diagnosis
first_ADHD <- data_flagged %>%
  arrange(PatientID, encounter_date) %>%
  filter(ADHD_dx == 1) %>%
  group_by(PatientID) %>%
  summarize(first_ADHD_date = min(encounter_date),
            age_at_first_ADHD_dx = AgeInYears[which.min(encounter_date)])

# Summarizing the rows (within Patient) where there's a MDD diagnosis
first_MDD <- data_flagged %>%
  arrange(PatientID, encounter_date) %>%
  filter(MDD_dx == 1) %>%
  group_by(PatientID) %>%
  summarize(first_MDD_date = min(encounter_date),
            age_at_first_MDD_dx = AgeInYears[which.min(encounter_date)])

################################################################################
# Merging the summary data together

patient_lev <- full_join(summary, first_ADHD, by = "PatientID")
patient_lev <- full_join(patient_lev, first_MDD, by = "PatientID")

################################################################################
# Re-joining with demographic information

# Making a separate dataframe for the demos
demos <- data_flagged[ ,c("PatientID", "GenderCode", "GenderName", "RaceCode")]

# This only keeps rows with distinct information (i.e., gets rid of all redundant duplicate info) 
demos <- demos %>%
  distinct()

# Joining demos back with the patient_level data
patient_lev <- full_join(demos, patient_lev)

```

# Categorizing
Categorizing patients have either having (1) or not having (1) any diagnosis for ADHD or MDD based on whether the first diagnosis date is NA or not
```{r}

patient_lev <- patient_lev %>%
  mutate(ADHD_ny = ifelse(is.na(first_ADHD_date), 0, 1),
         MDD_ny = ifelse(is.na(first_MDD_date), 0, 1))

# 4 patients w/ ADHD
table(patient_lev$ADHD_ny)

# 2 patients w/ MDD
table(patient_lev$MDD_ny)

```

# Saving patient_level summary data
```{r}

write.fst(patient_lev, "Example_pt_level_summary_data.fst")

```
