---
title: "3_Analyses"
author: "Emily Urban-Wojcik"
date: "2025-04-16"
output: html_document
---

# Note
This script provides a very simple structure for how you might start to look at prevelance estimates using DataDirect data. It begins by looking at descriptive statistics of the sample then moves to a formal chi-square analysis, which can tell you whether the proportion of individuals who are in one diagnosis category differs by whether or not they are also in the other diagnosis category. 

# Loading libraries/packages & paths
```{r}

# Un-comment and run the below lines to install these packages if you don't already have them
# install.packages("psych")
# install.packages("dplyr")
# install.packages("fst")
# install.packages("car")             
# install.packages("multcomp")        
# install.packages("ggplot2")
# install.packages("jtools")

# Opening the packages we need
library(psych)
library(dplyr)
library(fst)
library(ggplot2)
library(jtools)

# Set your working directory; Replace with your own path; This assumes you have separate directories for raw and cleaned data within your data directory
setwd("/Users/emurbanw/University of Michigan Dropbox/MED-EFDC-DDC-internal/Tools_to_Share/DataDirect/data/clean/")

# Opening data
patient_lev <- read.fst("Example_pt_level_summary_data.fst")

```

# Descriptives
```{r}

# 5 patients
length(unique(patient_lev$PatientID))

# ADHD + MDD  ADHD only 
#          2          2 
table(patient_lev$Group)

#    vars n mean    sd median trimmed  mad min max range skew kurtosis   se
# X1    1 4 36.5 19.67     31    36.5 12.6  20  64    44 0.49    -1.89 9.84
describe(patient_lev$age_at_first_ADHD_dx)

# Female   Male 
#   2       3  
table(patient_lev$GenderName)

# Female   Male 
#      1      3
table(patient_lev$GenderName[!is.na(patient_lev$Group)])

 # A AA  C 
 # 1  1  3
table(patient_lev$RaceCode)

```

# Chi-square
```{r}

# Counts/Percentages of patients in each Group that are Male vs. Female
table_data <- table(patient_lev$MDD_ny, patient_lev$ADHD_ny)
table_data
prop.table(table_data, margin = 2) * 100 # percentages within group

# Testing for group differences
chi2 <- table(patient_lev$MDD_ny, patient_lev$ADHD_ny)
chisq.test(chi2)

```